Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-05-02 Manuscript/manuscript.Rmd
Finish methods

## Computational framework for TAD boundary prediction using random forests

To make the data amenable for model training, the genome was first binned into equally sized intervals defined by the data resolution (*r*) of the corresponding contact matrix. The binning was initiated at the chromosome-specific genomic coordinate marking half of the resolution, and continued in congruent intervals until the end of the chromosome (*mod r* + *r/2*). This was to ensure that called (ground truth) TAD boundaries would be positioned in the center of respective genomic bins. Genomic bins were then defined as TAD boundary regions (*Y = 1*) if they contained a TAD boundary, and non-TAD boundary regions (*Y = 0*) otherwise, thus establishing the binary response vector (**Y**) used for random forest classification.

Cell type-specific genomic annotations were used to build the predictor space for of our random forest models. We created four types of predictors: *Peak Signal Strength* - [???include definition] (refered to hereafter as *Signal*), *Overlap Count (OC)* - the total number of overlaps between a bin and genomic annotation regions;  *Overlap Percent (OP)* - the percent of overlap between the bin and the total width of genomic annotation regions overlapping it; *Log2 Distance* - the $log_{2}$ of the distance in bases (refered to hereafter as *Distance*), from the center of the bin to the center of the nearest genomic annotation region.

Chromosome-specific datasets used for modeling were composed of two components: the response vector **Y**, an indicator variable separating between TAD and non-TAD boundary regions, and the predictor-type specific data matrix, **X** = {$X_1$, $X_2$, ..., $X_p$}, where $p$ was the set of genomic annotations. Each dataset was concatenated to form a whole genome dataset. The dataset was then split into a 7:3 ratio for training and testing respectively.

To assess the impact of class imbalance (CI), defined as the proportion of TAD boundary regions to non-TAD boundary regions, we evaluated three re-sampling techniques including *Random Over-Sampling (ROS)*, *Random Under-Sampling (RUS)*, and *Synthetic Minority Over-Sampling Technique (SMOTE)*. For ROS, the minority class was sampled with replacement to obtain the same number of data points in the majority class. For RUS, the majority class was sampled without replacement to obtain the same number of data points in the minority class. For SMOTE, under-sampling was performed without replacement from the majority class, while over-sampling was performed by creating new synthetic observations using the $k=5$ minority class nearest neighbors [@chawla2002smote] (implemented in the `DMwR` v.0.4.1 R package). We restricted the algorithm to 100% over-sampling and 200% under-sampling, which creates perfectly balanced classes.

@@

Random forest classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, predictor types, and re-sampling techniques. Three-fold cross validation was used to tune the number of annotations, *p*, to consider at each node. The number of trees that were aggregated for each random forest model was set to the default, $n=500$. The performance of each model was compared using Balanced Accuracy (BA), defined as

@@

## Feature reduction and predictive importance

Many genomic annotations, notably architectural proteins, tend to exhibit an extensive pattern of colocalization (correlation) [@van2014insulator]. To suitably reduce the feature space and prevent overfitting, we implemented recursive feature elimination to estimate the optimal number of necessary features to include in the best fitting random forest model. The best performing subset was identified, ranging from 2 to the maximum number of features by powers of 2. We then aggregated the predictive importance of the union of the optimal set of features across chromosomes using the mean decrease in node impurity among permuted features in out-of-bag samples to determine the most common and top ranked annotations for predicting TAD boundary regions.

## Precise TAD boundary prediction at bp resolution

Next, we aimed to investigate whether we could alleviate resolution limitations that plague conventional TAD calling algorithms. To do so, we extended our optimally built TAD prediction models to obtain predictions at base pair (bp) resolution. Naturally, when dealing with such fine resolution, there will exist collections of spatially close base pairs with similar predictor profiles (i.e. similar distances), and will thus yield similar predictions. To guard against inflating the number of predicted TAD boundaries we developed a technique that combines both density-based spatial clustering (DBSCAN) and partitioning techniques to identify more precise TAD boundaries throughout the linear genome.

We first established a *predicted TAD boundary area* (PTBA) as a collection of adjacent base pairs that have been identified as being a predicted TAD boundary with probability 1 according to the random forest model. We then implemented DBSCAN on the distance matrix (in base pairs) constructed from the midpoints of each PTBA using the R packacge *dbscan* (version 1.1-5) [@hahsler2019dbscan]. To identify the optimal number of clusters, *k*, the eps parameter ($\epsilon$) was assigned as the respective data resolution. The heuristic being that PTBA midpoints with genomic distances less than resolution should occupy the same designated cluster. The corresponding MinPts parameter was set to 3. We defined these *k* clusters of PTBAs as *predicted TAD boundary regions* (PTBR).

@@

To evaluate if our predicted TAD boundaries are biologically significant we asssessed the signal strength obtained from key ChIP-Seq peak elements known to be enriched at TAD boundaries including CTCF, RAD21, SMC3, and ZNF143. To do so we used enriched heatmaps and Wilcoxon Rank-Sum tests to compare distances from selected ChiP-seq data to our predicted TAD boundaries compared to ARROWHEAD boundaries.

Furthermore, we assessed the conservation of our predicted TAD boundaries across cell types and compared with ARROWHEAD. Boundaries were flanked by resolution and overlaps were evaluated visually using venn diagrams and Jaccard Indices. Jaccard is a metric used to measure overlap between sets defined as

@@

where A and B are genomic regions.

>>>>

## Computational framework for TAD boundary prediction

In Hi-C, each chromosome is binned into non-overlapping regions of the same length *r*, typically, 5kb and above. The *r* parameter defines the resolution of Hi-C data. Consequently, the location of TAD boundaries is limited by the resolution *r*. We designed a strategy called _shifted binning_ that partitions the genome into regions of the same length *r*, but with middle points corresponding to boundaries defined by the original binning.

To create shifted binning, the first chromosome-specific bin was set at half of the resolution *r*, and continued in congruent intervals of length *r* until the end of the chromosome (*mod r* + *r/2*). The shifted genomic bins, referred hereafted as bins for simplicity, were then defined as TAD boundary regions (*Y = 1*) if they contained a TAD boundary, and non-TAD boundary regions (*Y = 0*) otherwise, thus establishing the binary response vector (**Y**) used for classification (Figure 1).

Cell type-specific genomic annotations were used to build the predictor space for our TAD boundary prediction framework. Bins were annotated by the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), the average [??? Is that correct? Or, max?] signal strength of the corresponding annotation (*Peak Signal Strength*), and the $log_{2}$ of the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*).

Chromosome-specific datasets used for modeling were composed of two components: the response vector $Y_N$, an indicator variable separating between TAD and non-TAD boundary regions, and the predictor-type specific data matrix, $X = X_1, X_2, ..., X_p$ [??? Need to incorporate N, use proper math notation, maybe $\in$], where $n \in N$ is the number of bins (different for each chromosome) and $p$ is the number of genomic annotations (depends on the type of annotartions). Chromosome-specific dataseta were concatenated to form a whole genome dataset. The dataset was then split into a 7:3 ratio for training and testing, respectively.

## Class imbalance assessment

To assess the impact of class imbalance (CI), defined as the proportion of TAD boundary regions to non-TAD boundary regions, we evaluated three re-sampling techniques including *Random Over-Sampling (ROS)*, *Random Under-Sampling (RUS)*, and *Synthetic Minority Over-Sampling Technique (SMOTE)*. For ROS, the minority class was sampled with replacement to obtain the same number of data points in the majority class. For RUS, the majority class was sampled without replacement to obtain the same number of data points in the minority class. For SMOTE, under-sampling was performed without replacement from the majority class, while over-sampling was performed by creating new synthetic observations using the $k=5$ minority class nearest neighbors [@chawla2002smote] (implemented in the `DMwR` v.0.4.1 R package). We restricted the algorithm to 100% over-sampling and 200% under-sampling, which creates perfectly balanced classes. [??? Of what size?]

@@

Random forest classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, predictor types, and re-sampling techniques. Three-fold cross validation was used to tune the number of annotations, *p*, to consider at each node. The number of trees that were aggregated for each random forest model was set to the default, $n=500$. The performance of each model was compared using Balanced Accuracy (BA), defined as the average of sensitivity and specificity: [??? Correct?]

@@

## Feature selection and predictive importance

Many genomic annotations, notably architectural proteins, tend to exhibit an extensive pattern of colocalization (correlation) [@van2014insulator]. To suitably reduce the feature space and prevent overfitting, we implemented recursive feature elimination to estimate the optimal number of necessary features to include in the best fitting random forest model. The best performing subset was identified, ranging from 2 to the maximum number of features by power of 2. We then aggregated the predictive importance of the union of the optimal set of features across chromosomes using the mean decrease in node impurity among permuted features in out-of-bag samples to determine the most common and top ranked annotations for predicting TAD boundary regions.

## TAD boundary prediction at bp resolution

To investigate whether we could alleviate limitations of resolution *r* of Hi-C data that limits the precision of conventional TAD calling algorithms, we applied our optimally built TAD prediction models for predicting TAD boundaries at base pair (bp) resolution. [??? Write how your vector Y and annotation matrices change]. Naturally, when dealing with such fine resolution, there will be spatially close base pairs with similar predictor profiles (i.e. similar overlap counts, distances). To guard against inflating the number of predicted TAD boundaries we developed a technique that combines both density-based spatial clustering (DBSCAN) and partitioning techniques to identify more precise TAD boundaries throughout the linear genome.

We first established a *predicted TAD boundary area* (PTBA) as a collection of adjacent base pairs that have been identified as being a predicted TAD boundary with probability 1 according to the random forest model. We then implemented DBSCAN on the distance matrix (in base pairs) constructed from the midpoints of each PTBA using the R packacge *dbscan* (version 1.1-5) [@hahsler2019dbscan]. To identify the optimal number of clusters, *k*, the eps parameter ($\epsilon$) was assigned as the respective data resolution. The heuristic being that PTBA midpoints with genomic distances less than resolution should occupy the same designated cluster. The corresponding MinPts parameter was set to 3. We defined these *k* clusters of PTBAs as *predicted TAD boundary regions* (PTBR). [??? This deserves a schematic figure, before Figure 5. And part of it may be in Results] [??? I would switch the notation. First, you detect REGIONS. They can be adjacent, clustered together, forming AREA. Within this area, you can do pam clustering, and identify POINT]

@@

To evaluate biological significance of our predictions, we evaluated the average signal profile of CTCF, RAD21, SMC3, and ZNF143 centered on the predicted TAD boundaries. The signal proviles were plotted using the `EnrichedHeatmap` R package [??? Add version and citation]. Furthermore, we compared the [??? log2?] distance between TAD boundaries and CTCF, RAD21, SMC3, and ZNF143 binding sites using Wilcoxon Rank-Sum test.

Under the hypothesis that TAD boundaries are highly concerved across cell types [@dixon2012topological; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa], we assessed the conservation of our predicted TAD boundaries across cell types and prediction algorithms. Boundaries were flanked by resolution and overlaps were evaluated using Venn diagrams [??? Package?] and Jaccard index defined as

@@

where A and B are sets of genomic regions.
